---
---
title: "Ordinary Least Squares (OLS) Regression"
author: "Dr. Zahid Asghar"
format: revealjs
editor: visual
highlight-style: github
---

## What is OLS?

Ordinary Least Squares (OLS) is a method to estimate the coefficients of a linear regression model.

We assume the model:

$$
y_i = \beta_0 + \beta_1 x_i + \varepsilon_i
$$

Where:
- \( y_i \) is the dependent variable
- \( x_i \) is the independent variable
- \( \beta_0 \), \( \beta_1 \) are the coefficients
- \( \varepsilon_i \) is the error term

---

## Objective of OLS

OLS minimizes the **sum of squared residuals**:

$$
\text{Minimize} \quad \sum_{i=1}^n (y_i - \hat{y}_i)^2 = \sum_{i=1}^n (y_i - \beta_0 - \beta_1 x_i)^2
$$

The solution yields the best linear unbiased estimators (BLUE) under the Gauss-Markov assumptions.

---

## Estimating Coefficients

The OLS estimators are derived by solving:

\[
\begin{align*}
\hat{\beta}_1 &= \frac{\sum (x_i - \bar{x})(y_i - \bar{y})}{\sum (x_i - \bar{x})^2} \\
\hat{\beta}_0 &= \bar{y} - \hat{\beta}_1 \bar{x}
\end{align*}
\]

---

## OLS Assumptions

To ensure valid inference, OLS relies on several assumptions:

1. **Linearity**: Model is linear in parameters.
2. **Independence**: Observations are independent.
3. **Homoscedasticity**: Constant variance of residuals.
4. **Normality**: Residuals are normally distributed (for inference).
5. **No multicollinearity**: Predictors are not highly correlated.

---

## Matrix Formulation

For multiple regression, the model becomes:

$$
\mathbf{y} = \mathbf{X} \boldsymbol{\beta} + \boldsymbol{\varepsilon}
$$

- \( \mathbf{y} \): \( n \times 1 \) vector of responses  
- \( \mathbf{X} \): \( n \times p \) matrix of predictors (with intercept)  
- \( \boldsymbol{\beta} \): \( p \times 1 \) vector of coefficients  
- \( \boldsymbol{\varepsilon} \): error terms

Estimator:

$$
\hat{\boldsymbol{\beta}} = (\mathbf{X}^\top \mathbf{X})^{-1} \mathbf{X}^\top \mathbf{y}
$$

---

## OLS in R

```{r}
data(mtcars)
model <- lm(mpg ~ wt, data = mtcars)
summary(model)
```

---

## Model Visualization

```{r}
plot(mtcars$wt, mtcars$mpg, main = "MPG vs Weight", xlab = "Weight", ylab = "MPG")
abline(model, col = "blue")
```

---

## Diagnostics

Check model validity:

```{r}
par(mfrow = c(2,2))
plot(model)
```

Use these plots to assess:
- Residuals vs Fitted
- Normal Q-Q
- Scale-Location
- Residuals vs Leverage

---

## Conclusion

- OLS estimates the linear relationship between variables.
- Assumptions must be checked for valid inference.
- Extendable to multiple and polynomial regression.

---

```

---

### ✅ To use this:

1. Save it as `ols_lecture.qmd` in your RStudio project.
2. Click "Render" to generate the slides.
3. Math is auto-rendered using MathJax in RevealJS.

Would you like me to add speaker notes or make this version more interactive (like showing plots on click)?