---
title: "Farm to Failure: Higher Taxes Feeding Informal Sector"
subtitle: "Economic Policy Analysis"
author: "Zahid Asghar"
date: "May 19, 2025"
format: 
  revealjs:
    theme: simple
    slide-number: true
    logo: https://your-logo-here.png
    footer: "Source: Business Recorder"
---

## Executive Summary

:::: {.columns}

::: {.column width="60%"}
- Federal government's high taxes on documented sectors **fueling informal economy**
- Creating a **double negative impact**:
  - Reduced consumption of nutritional products
  - Growth of informal markets that evade taxes and food standards 
- Focus sectors: **fruit juices** and **dairy products** (including infant formula)
:::

::: {.column width="40%"}
![](https://via.placeholder.com/400x300)
:::

::::

## The Laffer Curve Effect

:::: {.columns}

::: {.column width="60%"}
- Higher tax rates becoming **counterproductive**
- Fruit beverage industry example:
  - FY20: 5% FED imposed → volumes ↓ 7%
  - FY23: 10% FED mid-year, 20% in FY24 → volumes ↓ 45%
  - Result: Government revenues now **falling**
:::

::: {.column width="40%"}
```{r}
#| echo: false
#| fig-cap: "Laffer Curve Illustration"

# Code for visualization would go here
```
:::

::::

## Impact on Fruit Juice Industry

- Combined tax rate now at **42%**
- Creates market opportunity for non-compliant players
- Consumers shifting to **lower-quality, untaxed products**
- Declining investment in value chain
- **Increasing fruit wastage** that could be processed
- Export potential diminishing

## Dairy Industry Challenges

:::: {.columns}

::: {.column width="55%"}
- Pakistan's packaged milk market still **nascent** (<10%)
  - Compared to Egypt (40%) and Turkey (60%)
- Shift from zero rating to **18% GST** in FY24
  - Result: Industry volumes ↓ 20%+
- Tax only on packaged products
- Even loose milk prices increased to maintain price difference
:::

::: {.column width="45%"}
```{r}
#| echo: false
#| fig-cap: "Packaged Milk Market Comparison"

# Code for visualization would go here
```
:::

::::

## Health Implications

- **Loose milk risks**: adulteration, carcinogenic substances
- Pasteurization solves majority of problems
- Most countries have low/no VAT on packaged dairy
- Critical for Pakistan where:
  - 70% of nutrition from wheat flour
  - Dairy provides affordable alternative protein
  - Two-thirds of packaged milk consumers are from **low-middle income groups**

## Infant Formula & Nutrition

:::: {.columns}

::: {.column width="60%"}
- Tax imposed on:
  - Infant formula
  - Baby food
  - Fortified nutrition products
- Problematic in a country with:
  - High maternal malnutrition
  - Prevalent child stunting
- While breastfeeding is optimal, formula necessary for mothers with deficiencies
:::

::: {.column width="40%"}
![](https://via.placeholder.com/400x300)
:::

::::

## Economic Losses

- **Value chain disruption** in both sectors
- Farmers losing income opportunities
- Increased food wastage
- Lost export potential:
  - Dairy industry: potential $150 million annual exports
  - Fruit juices: Already exported to 30 countries
  - Growing interest in Pakistani fruits and Halal products

## Policy Recommendations

1. **Revise tax strategy** on essential food items
2. Consider **differential tax rates** based on nutritional value
3. Focus on **enforcement of standards** rather than high taxation
4. Support **value chain development** in agricultural sectors
5. Promote **formalization through incentives** rather than punitive measures
6. Review tax policy on infant formula and nutritional products



**Questions & Discussion**

Source: "Farm to failure: higher taxes feeding informal sector"  
By Ali Khizar, Business Recorder  
Published May 19, 2025
