---
title: "Economy of Pakistan"
subtitle: "Measuring National Output: GDP, GNP & What the Numbers Hide"
author: "Zahid Asghar"
institute: "School of Economics, Quaid-i-Azam University, Islamabad"
date: "Spring 2026"
format:
  revealjs:
    theme: simple
    slide-number: true
    width: 1280
    height: 720
    margin: 0.07
    transition: slide
    background-transition: fade
    footer: "Economy of Pakistan · GDP & GNP · Zahid Asghar · QAU"
execute:
  echo: false
  warning: false
  message: false
---

```{=html}
<style>
@import url('https://fonts.googleapis.com/css2?family=Playfair+Display:ital,wght@0,700;1,400&family=Source+Sans+3:wght@300;400;600;700&family=JetBrains+Mono:wght@400;600&display=swap');

:root {
  --green:       #1a6b3c;
  --green-dark:  #0e2a1a;
  --green-light: #2d8f5a;
  --green-pale:  #e8f3ed;
  --gold:        #c9a84c;
  --gold-lt:     #f0d98a;
  --off-white:   #f6f8f3;
  --text-dark:   #0e1f13;

  /* was #3a5244 (too faint on projectors) */
  --text-muted:  #2b3f34;

  --red:         #c0392b;
  --amber:       #b45309;
  --amber-bg:    #fffbeb;
  --blue:        #1d5fa8;
  --blue-bg:     #eff6ff;
  --border:      #cce0d4;
}

.reveal {
  font-family: 'Source Sans 3', 'Segoe UI', sans-serif;
  font-size: 26px;
  color: var(--text-dark);
}
.reveal .slides { text-align: left; }

.reveal h1, .reveal h2, .reveal h3 {
  font-family: 'Playfair Display', Georgia, serif;
  text-transform: none;
  letter-spacing: -0.01em;
  line-height: 1.2;
}
.reveal h2 {
  font-size: 1.45em;
  color: var(--green);
  border-left: 5px solid var(--gold);
  padding-left: 0.5em;
  margin-bottom: 0.55em;
}
.reveal h3 { font-size: 1.05em; color: var(--green-dark); margin-bottom: 0.3em; }
.reveal p  { margin: 0.35em 0; line-height: 1.7; }
.reveal ul, .reveal ol { margin-left: 1.2em; line-height: 1.8; }
.reveal li { margin-bottom: 0.3em; }
.reveal strong { color: var(--green-dark); font-weight: 700; }

/* Default slide background */
.reveal .slide-background-content { background: var(--off-white); }

/* Dark slides */
.reveal .slides section[data-background-color="#0e2a1a"] h1,
.reveal .slides section[data-background-color="#0e2a1a"] h2,
.reveal .slides section[data-background-color="#0e2a1a"] h3,
.reveal .slides section[data-background-color="#0e2a1a"] p,
.reveal .slides section[data-background-color="#0e2a1a"] li { color: #fff !important; }
.reveal .slides section[data-background-color="#0e2a1a"] h2 {
  border-left-color: var(--gold) !important;
}
.reveal .slides section[data-background-color="#0e2a1a"] strong { color: var(--gold-lt) !important; }

.reveal .slides section[data-background-color="#1a2744"] h2,
.reveal .slides section[data-background-color="#1a2744"] h3,
.reveal .slides section[data-background-color="#1a2744"] p,
.reveal .slides section[data-background-color="#1a2744"] li { color: #fff !important; }
.reveal .slides section[data-background-color="#1a2744"] strong { color: var(--gold-lt) !important; }

/* Part-title slides */
.part-slide {
  display: flex;
  flex-direction: column;
  justify-content: center;
  align-items: flex-start;
  height: 68vh;
  padding: 0;
}
.part-label {
  font-family: 'JetBrains Mono', monospace;
  font-size: 0.68em;
  letter-spacing: 0.2em;
  text-transform: uppercase;
  color: var(--gold);
  margin-bottom: 0.5em;
}
.part-title {
  font-family: 'Playfair Display', Georgia, serif;
  font-size: 2.4em;
  color: #fff;
  line-height: 1.1;
  margin-bottom: 0.4em;
}
.part-sub { font-size: 1.0em; color: rgba(255,255,255,0.7); }

/* Boxes */
.box {
  padding: 0.9em 1.1em;
  border-radius: 0 8px 8px 0;
  font-size: 0.93em;
  line-height: 1.7;
  color: var(--text-dark);
}
.box.green   { background: var(--green-pale); border-left: 4px solid var(--green); }
.box.warning { background: #fdf2f2;           border-left: 4px solid var(--red); }
.box.info    { background: var(--blue-bg);    border-left: 4px solid var(--blue); }
.box.gold    { background: var(--amber-bg);   border-left: 4px solid #d97706; }
.box.dark    { background: var(--green-dark); border-left: 4px solid var(--gold);
               color: rgba(255,255,255,0.92); }
.box strong  { display: block; margin-bottom: 0.2em; }
.box.green strong  { color: var(--green); }
.box.warning strong { color: #991b1b; }
.box.info strong   { color: var(--blue); }
.box.gold strong   { color: #92400e; }
.box.dark strong   { color: var(--gold-lt); }

/* Eq box */
.eq-box {
  background: var(--green-dark);
  color: #fff;
  border-radius: 8px;
  padding: 0.65em 1.4em;
  text-align: center;
  font-size: 1.0em;
  margin: 0.4em 0 0.8em;
}

/* Def box */
.def-box {
  background: var(--green-pale);
  border-left: 5px solid var(--green);
  border-radius: 0 10px 10px 0;
  padding: 0.85em 1.2em;
  font-size: 1.0em;
  line-height: 1.7;
  margin-bottom: 0.9em;
}
.def-box strong { color: var(--green); }

/* Highlight banner */
.highlight-box {
  background: var(--green);
  color: #fff;
  border-radius: 8px;
  padding: 0.65em 1.2em;
  text-align: center;
  font-size: 0.95em;
  font-weight: 600;
  margin-top: 0.75em;
}
.highlight-box strong { color: var(--gold-lt) !important; font-weight: 700; }

/* Cycle */
.cycle-step {
  border-radius: 0 0 10px 10px;
  padding: 0.9em 0.7em;
  font-size: 0.86em;
  text-align: center;
  color: #fff;
  line-height: 1.7;
}
.cycle-step strong { color: #fff !important; display: block; margin-bottom: 0.25em; }
.step-red   { background: rgba(192,57,43,.28);  border-top: 3px solid var(--red); }
.step-amber { background: rgba(180,83,9,.28);   border-top: 3px solid #d97706; }
.step-green { background: rgba(45,143,90,.28);  border-top: 3px solid var(--green-light); }
.step-blue  { background: rgba(29,95,168,.28);  border-top: 3px solid var(--blue); }

.cycle-note {
  background: rgba(201,168,76,0.13);
  border: 1px solid rgba(201,168,76,0.35);
  border-radius: 8px;
  padding: 0.6em 1.1em;
  color: rgba(255,255,255,0.9);
  font-size: 0.88em;
  text-align: center;
  margin-top: 0.7em;
}
.cycle-note strong { color: var(--gold-lt) !important; }

/* Chart notes */
.chart-note       { font-size: 0.85em; text-align: center; color: var(--text-muted); margin-top: 0.3em; }
.chart-note-red   { font-size: 0.88em; text-align: center; color: var(--red); font-weight: 700; margin-top: 0.3em; }
.chart-note-white { font-size: 0.88em; text-align: center; color: rgba(255,255,255,0.88); margin-top: 0.3em; }

/* Tables */
.reveal table { font-size: 0.88em; border-collapse: collapse; width: 100%; }
.reveal table thead th {
  background: var(--green); color: #fff;
  padding: 0.5em 0.85em; text-align: left; font-size: 0.9em;
}
.reveal table tbody td { padding: 0.45em 0.85em; border-bottom: 1px solid var(--border); }
.reveal table tbody tr:nth-child(even) td { background: var(--green-pale); }

/* Footer & slide number */
.reveal .footer       { color: var(--text-muted) !important; font-size: 0.5em !important; opacity: 1 !important; }
.reveal .slide-number { background: transparent; color: var(--text-muted); font-size: 0.55em; opacity: 1 !important; }

/* =========================================================
   CONTRAST FIXES (KEEP AT END so these override everything)
   ========================================================= */

/* Ensure text never gets accidentally washed out */
.reveal p, .reveal li, .reveal td, .reveal th { opacity: 1 !important; }

/* Make footer + slide number readable on DARK slides */
.reveal .slides section[data-background-color="#0e2a1a"] .footer,
.reveal .slides section[data-background-color="#0e2a1a"] .slide-number,
.reveal .slides section[data-background-color="#1a2744"] .footer,
.reveal .slides section[data-background-color="#1a2744"] .slide-number {
  color: rgba(255,255,255,0.75) !important;
  opacity: 1 !important;
}

/* Chart note on light slides: slightly darker */
.chart-note { color: #1f3329 !important; opacity: 1 !important; }

/* Optional: disable “dim fragments before reveal” */
.reveal .slides section .fragment { opacity: 1 !important; }
.reveal .slides section .fragment.visible { opacity: 1 !important; }

/* Links: avoid washed out look */
.reveal a { color: var(--blue); text-decoration: underline; }
</style>
```

```{r setup, include=FALSE}
library(tidyverse)
library(knitr)
library(patchwork)
library(ggrepel)
```

---

## Economy of Pakistan {background-color="#0e2a1a"}

::: {style="font-family:'Playfair Display',Georgia,serif; font-size:1.2em; color:rgba(255,255,255,0.82); margin-top:0.3em; font-style:italic;"}
Measuring National Output: GDP, GNP & What the Numbers Hide
:::

::: {style="color:var(--gold-lt); font-size:0.88em; margin-top:1.4em; line-height:2.2;"}
Zahid Asghar  
School of Economics, QAU · Spring 2026
:::

---

## Learning Objectives {background-color="#0e2a1a"}

::: columns
::: {.column width="50%"}
- **Define** GDP and GNP — the exact conceptual difference
- **Apply** $GDP = C + I + G + (X-M)$ to Pakistan's real numbers
- **Explain** why Pakistan's GNP **exceeds** its GDP
:::
::: {.column width="50%"}
- **Evaluate** GDP as a welfare measure with Pakistan evidence
- **Compare** Pakistan with India, Bangladesh, Vietnam
- **Connect** measurement choices to policy decisions
:::
:::

---

## Part I — Gross Domestic Product {background-color="#0e2a1a"}

::: part-slide
::: part-label
Foundations
:::
::: part-title
Gross Domestic Product
:::
::: part-sub
Definition · Components · Pakistan's Numbers
:::
:::

---

## What is GDP? {background-color="#f6f8f3"}

::: def-box
**GDP** measures the monetary value of all **final** goods and services **produced within a country's geographical borders** during a specific time period.
:::

::: columns
::: {.column width="33%"}
::: {.box .green}
**① Final**

Count only the end product — the bread, not the wheat, flour, or dough. Each intermediate step would be counted twice otherwise.
:::
:::
::: {.column width="33%"}
::: {.box .gold}
**② Produced**

New output only. Not resale of existing assets, second-hand goods, or financial transactions.
:::
:::
::: {.column width="33%"}
::: {.box .info}
**③ Within Borders**

Territorial, not nationality-based. *Where* it is produced counts — not *who* produces it.
:::
:::
:::

---

## The Pakistan Illustration {background-color="#f6f8f3"}

::: columns
::: {.column width="50%"}
::: {.box .gold}
**🇨🇳 Chinese Engineer on CPEC (KP)**

Earns USD 8,000/month **inside Pakistan**

✅ **Adds to Pakistan's GDP**  
Production *within* Pakistan's borders

❌ **Does NOT add to Pakistan's GNP**  
Not a Pakistani resident
:::
:::
::: {.column width="50%"}
::: {.box .info}
**🇵🇰 Pakistani Doctor in the UK**

Earns £12,000/month **outside Pakistan**

❌ **Not in Pakistan's GDP**  
Production is outside Pakistan's borders

✅ **Adds to Pakistan's GNP**  
A Pakistani resident earning abroad
:::
:::
:::

::: {.box .green style="margin-top:0.8em;"}
This single contrast explains why **GNP > GDP** in Pakistan — and why that gap grows as emigration accelerates.
:::

---

## The Expenditure Approach {background-color="#f6f8f3"}

::: eq-box
$$GDP = C + I + G + (X - M)$$
:::

::: columns
::: {.column width="25%"}
::: {.box .warning}
**C — Consumption**
**~80% of GDP**

Highest in Asia. Saving rate only ~5%, leaving almost nothing for domestic investment.
:::
:::
::: {.column width="25%"}
::: {.box .warning}
**I — Investment**
**~13% of GDP**

A structural crisis. Bangladesh 32%, India 29%, Vietnam 27%. Pakistan cannot sustain 6%+ growth on this.
:::
:::
::: {.column width="25%"}
::: {.box .gold}
**G — Government**
**~20% of GDP**

Constrained by tax revenue of only ~10% of GDP. Over half goes to debt servicing.
:::
:::
::: {.column width="25%"}
::: {.box .warning}
**NX — Net Exports**
**Negative**

Imports persistently exceed exports. Pakistan runs a structural current-account deficit.
:::
:::
:::

---

## Pakistan's Investment Crisis {background-color="#1a2744"}

**Investment as % of GDP — Pakistan vs. Regional Peers (2024)**

```{r, fig.height=3.9, fig.width=11}
invest <- data.frame(
  country = c("Bangladesh","India","Vietnam","Malaysia","Pakistan"),
  inv     = c(32, 29, 27, 24, 13),
  type    = c("peer","peer","peer","peer","pak")
)
invest$country <- factor(invest$country, levels = invest$country)

ggplot(invest, aes(x = inv, y = country, fill = type)) +
  geom_col(width = 0.62) +
  geom_text(aes(label = paste0(inv, "%")),
            hjust = -0.2, color = "white", fontface = "bold", size = 6.5) +
  geom_vline(xintercept = 13, color = "#e8c547",
             linetype = "dashed", linewidth = 1.2) +
  annotate("text", x = 14, y = 5.45, label = "Pakistan 13%",
           color = "#e8c547", size = 5.5, hjust = 0, fontface = "bold") +
  scale_fill_manual(values = c("pak"="#c0392b","peer"="#2d8f5a"), guide="none") +
  scale_x_continuous(limits = c(0, 42)) +
  labs(x = NULL, y = NULL,
       caption = "Source: World Bank, ADB (2024). Pakistan: FY25 estimate.") +
  theme_void(base_size = 16) +
  theme(
    plot.background  = element_rect(fill = "#0e2a1a", colour = NA),
    panel.background = element_rect(fill = "#0e2a1a", colour = NA),
    axis.text.y = element_text(color = "white", size = 16, hjust = 1,
                               margin = margin(r = 10), face = "bold"),
    plot.caption = element_text(color = "#FFFFFF60", size = 10, hjust = 1,
                                margin = margin(t = 8))
  )
```

::: chart-note-white
Pakistan needs **25–30% investment/GDP** to sustain the **6–7% growth** needed to absorb 3–4 million new labour entrants each year.
:::

---

## Three Ways to Measure GDP {background-color="#f6f8f3"}

::: columns
::: {.column width="33%"}
::: {.box .green}
**📦 Expenditure** — *What was spent?*

$$C + I + G + NX$$

Counts final purchases across all agents. Most commonly reported by PBS, IMF, and ADB for Pakistan.
:::
:::
::: {.column width="33%"}
::: {.box .gold}
**🏭 Production** — *What was produced?*

Value added by sector:

- Agriculture **24%**
- Industry **19%**
- Services **57%**

Used for sectoral growth analysis.
:::
:::
::: {.column width="33%"}
::: {.box .info}
**💰 Income** — *What was earned?*

Wages + Profits + Rents + Interest

**Pakistan's problem:** 72% of non-agri workers are informal → wages unrecorded → output is systematically understated.
:::
:::
:::

::: {.box .green style="font-size:0.88em; margin-top:0.6em;"}
All three approaches yield the **same GDP** in theory. Large discrepancies reveal measurement failure — serious given Pakistan's ~$457B informal economy.
:::

---

## Pakistan's Sectoral Structure {background-color="#f6f8f3"}

```{r, fig.height=4.4, fig.width=11}
sectors <- data.frame(
  sector = c("Agriculture","Industry","Services"),
  share  = c(24.2, 19.1, 56.7),
  employ = c(38.5, 24.0, 37.5)
)
long_s <- sectors |>
  pivot_longer(c(share, employ), names_to = "type", values_to = "pct") |>
  mutate(
    type   = ifelse(type == "share", "GDP Share (%)","Employment Share (%)"),
    sector = factor(sector, levels = c("Agriculture","Industry","Services"))
  )
ggplot(long_s, aes(x = sector, y = pct, fill = type)) +
  geom_col(position = "dodge", width = 0.6) +
  geom_text(aes(label = paste0(pct, "%")),
            position = position_dodge(0.6), vjust = -0.4,
            fontface = "bold", size = 6.2) +
  scale_fill_manual(values = c("GDP Share (%)"="#1a6b3c",
                               "Employment Share (%)"="#c9a84c")) +
  scale_y_continuous(limits = c(0, 68),
                     labels = function(x) paste0(x, "%")) +
  labs(x = NULL, y = NULL, fill = NULL,
       title = "GDP Share vs. Employment Share — Pakistan FY2024–25",
       caption = "Source: Pakistan Bureau of Statistics (PBS)") +
  theme_minimal(base_size = 16) +
  theme(
    legend.position = "top",
    legend.text = element_text(size = 14),
    panel.grid.major.x = element_blank(),
    plot.title = element_text(face = "bold", color = "#0e2a1a", size = 16),
    plot.caption = element_text(color = "grey55", size = 10)
  )
```

::: chart-note-red
Agriculture employs **38.5%** of workers but generates only **24.2%** of GDP. Industry fell from ~24% (2005) → 19% today: **premature deindustrialisation**.
:::

---

## Real vs. Nominal GDP {background-color="#f6f8f3"}

::: eq-box
$$\text{Real GDP} = \frac{\text{Nominal GDP}}{\text{GDP Deflator}} \times 100$$
:::

::: columns
::: {.column width="48%"}

| Year | Inflation | Real Growth |
|:----:|:---------:|:-----------:|
| FY21 | 8.9% | 5.7% |
| FY22 | 12.2% | 6.1% |
| **FY23** | **29.2%** | **0.3%** |
| FY24 | 23.4% | 2.4% |
| FY25e | ~8.5% | ~3.0% |

:::
::: {.column width="52%"}
::: {.box .warning}
**⚠️ The FY23 Nominal Illusion**

Nominal GDP *appeared* to grow. With 29% inflation, real output grew just **0.3%** — essentially zero. Citizens reading nominal headlines were misled while poverty soared.
:::

::: {.box .gold style="margin-top:0.6em;"}
**FY25 "Recovery"**

Inflation fell from demand destruction and the global commodity supercycle reversing — not genuine supply-side improvement.
:::
:::
:::

---

## Part II — Gross National Product {background-color="#0e2a1a"}

::: part-slide
::: part-label
From Territory to Nationality
:::
::: part-title
Gross National Product
:::
::: part-sub
Why GNP > GDP in Pakistan — and what it reveals
:::
:::

---

## What is GNP? {background-color="#f6f8f3"}

::: def-box
**GNP / GNI** measures the **total income earned by a country's residents**, regardless of *where* that income is generated.
:::

::: eq-box
$$\text{GNP} = \text{GDP} + (\text{Factor Income Earned Abroad} - \text{Factor Income Paid to Foreigners})$$
:::

::: columns
::: {.column width="50%"}
::: {.box .green}
**➕ Adds to Pakistan's NFIA**

- Remittances from ~9M overseas Pakistanis
- Dividends from Pakistani investments abroad
- Salaries of Pakistanis in Gulf, UK, US
:::
:::
::: {.column width="50%"}
::: {.box .warning}
**➖ Reduces Pakistan's NFIA**

- Profits repatriated by MNCs in Pakistan
- Salaries of CPEC engineers sent to China
- Debt interest paid to foreign creditors (~$5B/yr)
:::
:::
:::

---

## Pakistan's Remittance Lifeline {background-color="#1a2744"}

**Remittance Inflows — USD Billion and % of GDP (FY20–FY25)**

```{r, fig.height=3.9, fig.width=11}
remit <- data.frame(
  year = c("FY20","FY21","FY22","FY23","FY24","FY25e"),
  usd  = c(23.1, 29.4, 31.2, 27.0, 30.3, 33.5),
  pct  = c(7.9, 9.8, 9.5, 8.8, 9.4, 9.8)
)
ggplot(remit, aes(x = year)) +
  geom_col(aes(y = usd), fill = "#2d8f5a", alpha = 0.9, width = 0.55) +
  geom_line(aes(y = pct * 3.4, group = 1),
            color = "#e8c547", linewidth = 2.0) +
  geom_point(aes(y = pct * 3.4), color = "#e8c547", size = 5) +
  geom_text(aes(y = usd + 1.5, label = paste0("$", usd, "B")),
            color = "white", fontface = "bold", size = 5.5) +
  geom_text(aes(y = pct * 3.4 + 1.8, label = paste0(pct, "%")),
            color = "#e8c547", fontface = "bold", size = 4.8) +
  scale_y_continuous(
    name = "USD Billion",
    sec.axis = sec_axis(~ . / 3.4, name = "% of GDP",
                        labels = function(x) paste0(x, "%"))
  ) +
  scale_x_discrete(expand = c(0.1, 0.1)) +
  labs(x = NULL,
       caption = "Source: State Bank of Pakistan (SBP); World Bank. FY25 = estimate.") +
  theme_minimal(base_size = 15) +
  theme(
    plot.background  = element_rect(fill = "#0e2a1a", colour = NA),
    panel.background = element_rect(fill = "#0e2a1a", colour = NA),
    axis.text   = element_text(color = "#FFFFFFD9", size = 13),
    axis.title  = element_text(color = "#c9a84c", size = 12),
    panel.grid  = element_line(color = "#FFFFFF14"),
    plot.caption = element_text(color = "#FFFFFF55", size = 10, hjust = 1)
  )
```

::: chart-note-white
**~9.4% of GDP** — one of the highest ratios globally. India: 3.5%. A **structural crutch**: reflects failure to create competitive domestic employment.
:::

---

## GDP vs. GNP: Key Differences {background-color="#f6f8f3"}

| | GDP | GNP / GNI |
|:--|:--|:--|
| **Core concept** | Territorial — *where* production occurs | Nationality — *who* earns the income |
| Chinese CPEC engineer in KP | ✅ In Pakistan's GDP | ❌ Not in Pakistan's GNP |
| Pakistani doctor in UK | ❌ Not in Pakistan's GDP | ✅ In Pakistan's GNP |
| Remittances | ❌ Not included | ✅ Core component |
| IMF programme targets | ✅ GDP-based | — |
| World Bank poverty measure | — | ✅ GNI per capita used |

::: highlight-box
Pakistan's GNP exceeds GDP by **~USD 25–30 billion** annually — remittances ($30B) far exceed profit repatriation by foreign firms (~$4B).
:::

---

## GNP Reconciliation — Pakistan FY24 {background-color="#f6f8f3"}

::: columns
::: {.column width="52%"}

| Item | USD Billion |
|:-----|------------:|
| GDP at market prices (FY24) | 374 |
| ＋ Remittances & factor income | +30.3 |
| ＋ Investment income abroad | +1.2 |
| － Profits repatriated by MNCs | −4.1 |
| － Debt interest paid abroad | −2.8 |
| **≈ GNP / GNI** | **≈ 399** |

:::
::: {.column width="48%"}
::: {.box .gold}
**⚠️ The Remittance Paradox**

High remittances make per-capita GNP > per-capita GDP — citizens appear *better off* than domestic production implies.

But every skilled Pakistani abroad is a **productivity loss** to the domestic economy. The premium compensates for structural failure — not genuine prosperity.
:::
:::
:::

---

## Part III — Regional Comparison {background-color="#0e2a1a"}

::: part-slide
::: part-label
Pakistan in Perspective
:::
::: part-title
Regional Comparison
:::
::: part-sub
What the numbers look like next to our neighbours
:::
:::

---

## Pakistan vs. Regional Peers {background-color="#f6f8f3"}

```{r, fig.height=4.5, fig.width=11}
peers <- data.frame(
  country   = c("Malaysia","Vietnam","India","Bangladesh","Pakistan"),
  gdp_pc    = c(13500, 4300, 2730, 2700, 1600),
  exports   = c(70, 87, 22, 15, 10),
  highlight = c("peer","peer","peer","peer","pak")
)
p1 <- ggplot(peers,
             aes(y = reorder(country, gdp_pc), x = gdp_pc, fill = highlight)) +
  geom_col(width = 0.65) +
  geom_text(aes(label = paste0("$", format(gdp_pc, big.mark=","))),
            hjust = -0.12, size = 5.5, fontface = "bold", color = "#0e2a1a") +
  scale_fill_manual(values = c("pak"="#c0392b","peer"="#2d8f5a"), guide="none") +
  scale_x_continuous(limits = c(0, 19000),
                     labels = function(x) paste0("$",x/1000,"k")) +
  labs(title = "GDP Per Capita (USD, 2024)", x = NULL, y = NULL) +
  theme_minimal(base_size = 14) +
  theme(panel.grid.major.y = element_blank(),
        axis.text = element_text(size = 13),
        plot.title = element_text(face = "bold", size = 14, color = "#0e2a1a"))

p2 <- ggplot(peers,
             aes(y = reorder(country, exports), x = exports, fill = highlight)) +
  geom_col(width = 0.65) +
  geom_text(aes(label = paste0(exports, "%")),
            hjust = -0.12, size = 5.5, fontface = "bold", color = "#0e2a1a") +
  scale_fill_manual(values = c("pak"="#c0392b","peer"="#2d8f5a"), guide="none") +
  scale_x_continuous(limits = c(0, 105),
                     labels = function(x) paste0(x, "%")) +
  labs(title = "Exports / GDP (2024)", x = NULL, y = NULL) +
  theme_minimal(base_size = 14) +
  theme(panel.grid.major.y = element_blank(),
        axis.text = element_text(size = 13),
        plot.title = element_text(face = "bold", size = 14, color = "#0e2a1a"))

p1 + p2
```

::: chart-note-red
Pakistan: **lowest exports/GDP (10%)** in the region · GDP per capita well below every peer
:::

---

## The Bangladesh Inflection Point {background-color="#1a2744"}

```{r, fig.height=4.5, fig.width=11}
diverge <- data.frame(
  year    = rep(c(2000,2005,2010,2015,2019,2022,2024), 2),
  country = rep(c("Bangladesh","Pakistan"), each = 7),
  gdp_pc  = c(420, 560, 750, 1080, 1830, 2460, 2700,
              610, 870, 1100, 1430, 1530, 1580, 1600)
)
ggplot(diverge, aes(x = year, y = gdp_pc, color = country, group = country)) +
  geom_line(linewidth = 2.2) +
  geom_point(size = 4.5) +
  geom_vline(xintercept = 2019, color = "#e8c547",
             linetype = "dashed", linewidth = 1.1) +
  annotate("text", x = 2019.5, y = 2400, label = "Crossover ~2019",
           color = "#e8c547", size = 5.5, hjust = 0, fontface = "bold") +
  scale_color_manual(values = c("Bangladesh"="#2d8f5a","Pakistan"="#c0392b")) +
  scale_y_continuous(labels = function(x) paste0("$", x),
                     breaks = seq(0, 3000, 500)) +
  scale_x_continuous(breaks = c(2000,2005,2010,2015,2019,2022,2024)) +
  labs(x = NULL, y = "GDP Per Capita (USD)", color = NULL,
       title = "GDP Per Capita: Pakistan vs. Bangladesh, 2000–2024",
       caption = "Source: World Bank WDI") +
  theme_minimal(base_size = 15) +
  theme(
    plot.background  = element_rect(fill = "#0e2a1a", colour = NA),
    panel.background = element_rect(fill = "#0e2a1a", colour = NA),
    plot.title   = element_text(color = "white", face = "bold", size = 15),
    axis.text    = element_text(color = "#FFFFFFD9", size = 13),
    axis.title   = element_text(color = "#c9a84c", size = 13),
    legend.text  = element_text(color = "white", size = 13),
    panel.grid   = element_line(color = "#FFFFFF14"),
    plot.caption = element_text(color = "#FFFFFF55", size = 10)
  )
```

::: chart-note-white
Bangladesh exports: **$6B → $55B** (2000–2024) · Pakistan: $9B → $30B · Female LFPR: **43% vs. 24%**
:::

---

## Part IV — What GDP Misses {background-color="#0e2a1a"}

::: part-slide
::: part-label
Critical Perspective
:::
::: part-title
What GDP Misses
:::
::: part-sub
Limitations · Pakistan Evidence · Alternative Measures
:::
:::

---

## GDP Limitations 1 & 2 {background-color="#f6f8f3"}

::: columns
::: {.column width="50%"}
::: {.box .warning}
**1. Ignores Income Distribution**

FY22: GDP grew **6.1%** — yet poverty *rose*.

GDP measures the size of the pie, not who gets a slice. Pakistan's growth has bypassed the bottom 40%.

World Bank (2025): poverty **45%**, extreme poverty up from 4.9% → **16.5%** despite positive GDP growth.
:::
:::
::: {.column width="50%"}
::: {.box .warning}
**2. Misses the Informal Economy**

**72%** of non-agricultural workers are informal.

The informal economy (~**$457B**) is *larger* than the formal economy (~$340B).

GDP understates actual output and inflates the apparent tax-to-GDP ratio.
:::
:::
:::

---

## GDP Limitations 3 & 4 {background-color="#f6f8f3"}

::: columns
::: {.column width="50%"}
::: {.box .gold}
**3. Excludes Non-Market Work**

Pakistan's female LFPR is just **~24%** — lowest in South Asia.

Millions of hours of domestic labour, childcare, and subsistence farming go entirely **uncounted** in GDP — a massive share of total welfare-generating activity.
:::
:::
::: {.column width="50%"}
::: {.box .gold}
**4. Ignores Environmental Costs**

The 2022 floods cost Pakistan an estimated **USD 30 billion** — roughly two years of GDP gains.

The destruction was **not subtracted** from GDP. Reconstruction spending actually *added* to it. True welfare progress is far slower than headlines suggest.
:::
:::
:::

---

## GDP Limitations 5 & 6 {background-color="#f6f8f3"}

::: columns
::: {.column width="50%"}
::: {.box .info}
**5. Ignores Quality of Life**

Pakistan ranks **164th / 193** on the UN Human Development Index (2024).

Life expectancy: **66 years** · Learning poverty among highest in Asia.

GDP growth has not translated into health or education gains — output and welfare have **decoupled**.
:::
:::
::: {.column width="50%"}
::: {.box .info}
**6. Quantity Not Quality of Growth**

A construction boom and a manufacturing productivity surge look **identical** in GDP figures.

Pakistan's recent growth: low-productivity services, remittance-fuelled consumption, and construction — **not** high-value manufacturing or technology.
:::
:::
:::

---

## Pakistan's GDP vs. HDI {background-color="#f6f8f3"}

```{r, fig.height=4.5, fig.width=11}
hdi_data <- data.frame(
  country   = c("Malaysia","Vietnam","India","Bangladesh","Pakistan","Nepal"),
  gdp_pc    = c(13500, 4300, 2730, 2700, 1600, 1340),
  hdi       = c(0.807, 0.726, 0.644, 0.661, 0.544, 0.601),
  highlight = c("peer","peer","peer","peer","pak","peer")
)
ggplot(hdi_data,
       aes(x = gdp_pc, y = hdi, color = highlight, label = country)) +
  geom_smooth(method = "lm", se = FALSE, color = "grey80",
              linetype = "dashed", linewidth = 1.1) +
  geom_point(aes(size = highlight), alpha = 0.92) +
  geom_text_repel(size = 5.8, fontface = "bold",
                  box.padding = 0.7, color = "#0e2a1a") +
  scale_color_manual(values = c("pak"="#c0392b","peer"="#1a6b3c"), guide="none") +
  scale_size_manual(values = c("pak"=9,"peer"=6), guide="none") +
  scale_x_continuous(
    labels = function(x) paste0("$", format(x, big.mark=",")),
    limits = c(1000, 16000)
  ) +
  scale_y_continuous(limits = c(0.50, 0.86)) +
  labs(x = "GDP Per Capita (USD, 2024)", y = "Human Development Index (2024)",
       title = "GDP Per Capita vs. HDI — Pakistan Punches Well Below Its Weight",
       caption = "Source: World Bank WDI; UNDP HDR 2024") +
  theme_minimal(base_size = 16) +
  theme(
    plot.title   = element_text(face = "bold", size = 16, color = "#0e2a1a"),
    axis.text    = element_text(size = 14),
    axis.title   = element_text(size = 13),
    plot.caption = element_text(color = "grey55", size = 11)
  )
```

::: chart-note-red
**Nepal** has a lower GDP per capita yet a **higher HDI (0.601 vs. 0.544)**. GDP is not converting into welfare.
:::

---

## Alternative Welfare Measures {background-color="#f6f8f3"}

::: columns
::: {.column width="50%"}
::: {.box .green}
**📊 HDI (UNDP)**  
Life expectancy + Education + GNI per capita  
Pakistan: **0.544 · Rank 164/193** (2024)
:::

::: {.box .green style="margin-top:0.75em;"}
**🔢 MPI (OPHI)**  
Poverty across health, education & living standards  
Pakistan: **38.4% multidimensionally poor** (2023)
:::
:::
::: {.column width="50%"}
::: {.box .green}
**💸 World Bank Poverty Rate**  
Updated poverty lines (HIES 2018–19 data)  
Pakistan: **45% poor · 16.5% extreme poverty** (2025)
:::

::: {.box .green style="margin-top:0.75em;"}
**🌿 Genuine Progress Indicator (GPI)**  
GDP adjusted for inequality, environment & household work  
No official GPI for Pakistan — but would be **far below** its GDP given flood losses and high inequality
:::
:::
:::

---

## Part V — Pakistan's Growth Problem {background-color="#0e2a1a"}

::: part-slide
::: part-label
Macro Diagnosis
:::
::: part-title
Pakistan's Growth Problem
:::
::: part-sub
The boom-bust cycle · What 6–7% growth actually requires
:::
:::

---

## How Much Growth Does Pakistan Need? {background-color="#f6f8f3"}

::: columns
::: {.column width="48%"}
**The arithmetic:**

- Population growth: **~1.9% per year**
- New labour entrants: **3–4 million per year**
- To hold unemployment steady → need **6–7% GDP growth**
- To reduce poverty meaningfully → sustained growth above 7%

**ADB FY26 forecast: 3.0%**  
**Government target: 4.2%**

::: {.box .warning style="margin-top:0.7em;"}
Both figures are **below the minimum** needed to prevent unemployment rising. Pakistan is falling further behind — not catching up.
:::
:::
::: {.column width="52%"}
::: {.box .dark}
**What 6–7% growth requires:**

🏭 Investment → **25–30% of GDP** *(currently 13%)*

📦 Exports diversified beyond textiles *(stuck at ~$30B)*

⚡ Energy → competitive industrial tariffs *(region's highest)*

📋 Tax base → **15–16% of GDP** *(currently ~10%)*

👩‍🎓 Human capital → reverse learning poverty *(generational commitment)*
:::
:::
:::

---

## Pakistan's 25-Programme IMF Cycle {background-color="#1a2744"}

Pakistan has engaged with the IMF **25 times since 1950** — roughly every 2–3 years on average.

::: columns
::: {.column width="20%"}
::: {.cycle-step .step-red}
**① Crisis**  
BOP collapse  
Reserves depleted  
Rupee crash  
Inflation spikes
:::
:::
::: {.column width="20%"}
::: {.cycle-step .step-amber}
**② IMF Entry**  
Austerity  
Rate hikes  
Devaluation  
Import squeeze
:::
:::
::: {.column width="20%"}
::: {.cycle-step .step-green}
**③ Stabilisation**  
Inflation falls  
Reserves rebuild  
Surplus achieved  
GDP recovers
:::
:::
::: {.column width="20%"}
::: {.cycle-step .step-blue}
**④ Premature Boom**  
Fiscal slippage  
Import surge  
Reform fatigue  
Programme exits
:::
:::
::: {.column width="20%"}
::: {.cycle-step .step-red}
**⑤ Next Crisis**  
Structural drivers  
unchanged  
↺ Return to ①
:::
:::
:::

::: cycle-note
Each stabilisation restores macro indicators — but never addresses the structural causes: **low exports · narrow tax base · energy circular debt · institutional fragility**
:::

---

## Cyclical or Structural? FY25 {background-color="#f6f8f3"}

::: columns
::: {.column width="50%"}
::: {.box .green}
**Government's FY25 Narrative ✅**

Inflation: 38% peak → **~8.5%**  
Primary fiscal surplus: **3 consecutive years**  
SBP reserves: rebuilt to **$14.5B**  
Policy rate: cut **1,100 bps** to 12%  
ADB growth forecast: **3.0%**
:::
:::
::: {.column width="50%"}
::: {.box .warning}
**What the Data Shows ⚠️**

Inflation fell from **demand destruction** — not supply reform  
Fiscal surplus via **crushing thin tax base** — not broadening it  
Reserves rebuilt by **suppressing imports** — not earning through exports  
Net FDI: **>80% retained earnings** — not fresh capital  
3.0% growth is a **"dead cat bounce"** — far below 6%
:::
:::
:::

::: highlight-box
The improvements are **cyclical, not structural**. The next crisis is a question of *when*, not *whether*.
:::

---

## Key Takeaways {background-color="#0e2a1a"}

::: columns
::: {.column width="50%"}
**On Measurement**

- GDP = **territorial** output; GNP = **resident** income
- Pakistan's **GNP > GDP** by ~$25–30B — remittances dwarf profit repatriation
- Use **real GDP** for welfare comparisons — nominal misled catastrophically in FY23
- Official GDP **understates** true activity — informal sector exceeds the formal
:::
::: {.column width="50%"}
**On Pakistan's Economy**

- Investment at **13% of GDP** is a structural crisis, not a cyclical dip
- Exports at **10% of GDP** — lowest in the region
- HDI rank **164/193** — far below what income level predicts
- **25 IMF programmes** since 1950 = stabilisation without structural transformation
:::
:::

---

## Conclusion {background-color="#0e2a1a"}

::: {style="font-family:'Playfair Display',Georgia,serif; font-size:1.2em; color:#f0d98a; margin-bottom:0.7em; font-style:italic;"}
"GDP is the most powerful single number we have — and among the most dangerous if used alone."
:::

GDP captures market-based economic activity consistently and comparably. For Pakistan it tracks the business cycle, validates IMF targets, and anchors budget planning.

But taken alone it **hides** 72% informal employment, **misses** USD 457B of uncounted activity, **ignores** the distribution that left 45% in poverty, and **fails** to subtract the $30B cost of the 2022 floods.

Pakistan's challenge is not just to grow GDP — it is to ensure growth is **investment-led, export-driven, and human-capital-intensive** enough to break the 25-programme IMF cycle and lift nearly half the population out of poverty.

---

## {background-color="#0e2a1a"}

::: {style="text-align:center; padding-top:1.5em;"}

::: {style="font-family:'Playfair Display',Georgia,serif; font-size:2.5em; color:#c9a84c; margin-bottom:0.7em;"}
Questions?
:::

::: {style="color:rgba(255,255,255,0.75); font-size:0.9em; line-height:2.4;"}
Zahid Asghar  
School of Economics, Quaid-i-Azam University  
zahid.asghar@qau.edu.pk
:::

::: {style="color:rgba(255,255,255,0.48); font-size:0.82em; margin-top:1em;"}
**Next Lecture:** National Income Accounting — The Circular Flow and Pakistan's Leakages
:::

:::
